---
title: 'Final Project: Child Welfare in the US'
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    theme: lumen
    vertical_layout: fill
runtime: shiny
resource_files:
- FinalDataforSubmission.csv

---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(plotly)
library(shiny)
library(waffle)
library(rsconnect)
library(scales)
library(packcircles)
library(treemapify)

# read in data 
childhealthdata <- read_csv("FinalDataforSubmission.csv")


#adding filtered variables 


MissedNeededCare <- childhealthdata %>%
  filter(ForegoneCare == "Yes")

uninsured <- childhealthdata %>%
  filter(InsGap_1920 == "Periods Uninsured")

childhealthdata <- childhealthdata %>% mutate(ACEs = recode_factor(ACEct_1920,
  "0" = "0",
  "1" = "1",
  "2" = "2",
  "3" = "3",
  "4" = "4",
  "5" = "5",
  "6" = "6", 
  "7" = "7",
  "8" = "8",
  "9" = "9",
  "10" = "10",
  "99" = "Missing",
  .ordered = TRUE,
  .default = "Unknown"))

childhealthdata <- childhealthdata %>% mutate(povlevel = recode_factor(povlev4_1920, 
  "1" = "Below 100% FPL",
  "2" = "100-199% FPL",
  "3" = "200-399% FPL",
  "4" = "Greater than 400% FPL",
  .ordered = TRUE,
  .default = "Unknown"))

childhealthdata <- childhealthdata %>% mutate(RaceCat = recode_factor(Race, 
  "1" = "Hispanic",
  "2" = "White",
  "4" = "Asian",
  "3" = "Black",
  "5" = "Other",
  "99" = "Unknown",
  .ordered = TRUE,
  .default = "Unknown"))
```

Demographics {data-orientation=columns}
===================================== 
About this Project {.sidebar}
--------------------------------------------------------------
This project was created for submission as a part of Rutgers Data Visualization course. It focuses on child wellbeing in the United States, as evidenced by the National Survey of Children's Health, 2019-2020. 

This dashboard displays demographics of children surveyed, as trends related to insurance status, receiving care, poverty, and Adverse Childhood Experiences (ACE).  

Created by [Rebecca Connelly]

Column {data-width=450} 
-------------------------------------------------------------

### Racial Distribution

```{r race}
#race 

RaceCount <- childhealthdata %>% 
  group_by(RaceCat) %>% 
  count() %>% arrange(desc(n)) %>%
  rename("children" = n)

RaceCount <- RaceCount %>% filter(!(RaceCat %in% "Unknown")) 

# race tree map 

RaceTreeMap <- ggplot(RaceCount, aes(area = children, fill = RaceCat, label = RaceCat)) +
  geom_treemap() + 
  scale_fill_viridis_d(option = "viridis", 
 direction = 1) +
   geom_treemap_text(fontface = "italic", colour = "white", place = "centre", min.size = 2, grow = TRUE)

RaceTreeMap



```


### Insurance Coverage
```{r insurance pie chart}

InsStatus <- childhealthdata %>% 
  group_by(InsGap_1920) %>% 
  count() %>% arrange(desc(n)) %>%
  rename("children" = n)
InsStatus <- InsStatus %>% filter(InsGap_1920 != "Unknown")

InsStatus$percentage <- InsStatus$children/sum(InsStatus$children)*100
InsStatus$Round_off <- round(InsStatus$percentage) 

#in plotly

piecolors <- c('rgb(68, 1, 84)', 'rgb(253, 231, 37)')

InsurancePieChart <- plot_ly(InsStatus, labels = ~InsGap_1920, values = ~children, type = 'pie', 
        textposition = 'outside',
        textinfo = 'percent',
        insidetextfont = list(color = '#FFFFFF'),
        hoverinfo = 'text',
        text = paste("Children: ", InsStatus$children),
        marker = list(colors = piecolors),
                      line = list(color = '#FFFFFF', width = 1),
        showlegend = TRUE)

InsurancePieChart <- InsurancePieChart %>% layout(title = 'Insurance Coverage',
      xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
      yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))

InsurancePieChart

```

Column {data-width=650}
-------------------------------------------------------------

### Age Distribution

```{r child age}

childhealthdata <- childhealthdata %>% rename(Age = Child.Age)

ageplot <- ggplot(childhealthdata) +
  aes(x = Age, text = paste0("Age: ", x, "<br> Children: ", ..count..)) +
  geom_histogram(bins = 35L, fill = "#21918c") +
  labs(
    x = "Age of Child",
    y = "Number of Children",
    title = "Age Distribution of Children",
  ) +
  coord_flip() +
    theme_minimal()

ggplotly(ageplot, tooltip = "text")
  

```

### Poverty Levels 

```{r poverty level chart}
povplot <- ggplot(childhealthdata) +
  aes(x = povlevel, text = paste0("Children: ", ..count.., "<br> Poverty Level: ", x)) +
  geom_bar(fill = "#440154") +
  labs(
    x = "Relation to Federal Poverty Level",
    y = "Number of Children",
  title = "Poverty Level of Children Sampled") +
    theme_minimal()


ggplotly(povplot, tooltip = "text")


```


Access 
=====================================

Row 1
-----------------------------------------------------------------------

### Uninsured Stats

```{r value box uninsured}

valueBox("6%", caption = "Percent Uninsured", icon = NULL, color = "#21918c", href = NULL)

```

### Poverty Stats

```{r value box poverty}
valueBox("28%", caption = "Percent in Poverty", icon = NULL, color = "#21918c", href = NULL)
```

### Needed Treatment Stats

```{r value box treatment }
valueBox("3%", caption = "Percent Who Need Treatment & Did Not Receive It", icon = NULL, color = "#21918c", href = NULL)
```


Row 2
-------------------------------------------------------------

### Care Not Received

```{r reasons care not received }
numineligible <- childhealthdata %>% filter(ineligible_1920 == 1) %>% count()

numunavailable <- childhealthdata %>% filter(unavailable_1920 == 1) %>% count()

numnoappt <- childhealthdata %>% filter(appointment_1920 == 1) %>% count()

numntransport <- childhealthdata %>% filter(transport_1920 == 1) %>% count()

numNotOpen <- childhealthdata %>% filter(NotOpen_1920 == 1) %>% count()

numCostIssue <- childhealthdata %>% filter(CostIssue_1920 == 1) %>% count()

Reason <- c("Not Eligible", "Unavailable in Area", "Couldn't Get Appointment", "Transportation or Childcare Issues", "Office Wasn't Open", "Cost")
reasonscount <- c(518, 602, 1129, 230, 596, 1118)

reasonsnocaredata <- data.frame(Reason, reasonscount)

reasonsnocareplot <- ggplot(reasonsnocaredata) +
  aes(x = Reason, y = reasonscount, text = paste0("Children: ", reasonscount, "<br> Reason: ", Reason)) +
  geom_bar(stat = "identity", fill = "#440154") +
  labs(
    x = "",
    y = "Number of Children",
    title = "Reasons for Missing Needed Care") +
  coord_flip() +
    theme_minimal()

ggplotly(reasonsnocareplot, tooltip = "text")

```

### Insurance & Poverty 

```{r insurance and poverty}

#plot showing over 50% of poor are uninsured 

insurancepovchart <- childhealthdata %>%
 filter(!(InsGap_1920 %in% "Unknown")) %>%
 ggplot() +
 aes(x = InsGap_1920, fill = povlevelgrouped, text = paste("Children: ", ..count.., "<br> Poverty Level: ", fill)) +
 geom_bar(position = "fill") +
 scale_fill_viridis_d(option = "viridis", 
 direction = 1) +
  scale_y_continuous(labels = scales::percent) +
 labs(x = "", y = "Percentage", title = "Insurance Status and Poverty Level in US",
 caption = "Based on National Survey of Children's Health, 2019-2020", fill = "Poverty Level") +
 coord_flip() +
 theme_minimal()

ggplotly(insurancepovchart, tooltip = "text")
```


Row {.tabset}
-------------------------------------------------------------

### Needed Care By Insurance Status

```{r care insurance}
#insured status by received care
CareInsChart <- childhealthdata %>%
 filter(!(InsGap_1920 %in% "Unknown")) %>%
 filter(!(ForegoneCare %in% "Unknown")) %>%
 ggplot() +
 aes(x = ForegoneCare, fill = InsGap_1920, text = paste("Children: ", ..count.., "<br> Insurance Status: ", fill)) +
 geom_bar(position = "fill") +
 scale_fill_viridis_d(option = "viridis", direction = 1) +
  scale_y_continuous(labels = scales::percent) +
 labs(x = "", 
 y = "Proportion of Children", title = "Children Receiving Necessary Care by Insurance Status", caption = "Based on National Survey of Children's Health, 2019-2020", 
 fill = "Insured Status") +
  coord_flip() +
 theme_minimal()


ggplotly(CareInsChart, tooltip = "text")
```

### By Poverty Level 

```{r care by poverty}
CarePovChart <- childhealthdata %>%
 filter(!(ForegoneCare %in% "Unknown")) %>%
 ggplot() +
 aes(x = ForegoneCare, fill = povlevelgrouped, text = paste("Children: ", ..count.., "<br> Poverty Level: ", fill)) +
 geom_bar(position = "fill") +
 scale_fill_viridis_d(option = "viridis", 
 direction = 1) +
  scale_y_continuous(labels = scales::percent) +
 labs(x = "", y = "Proportion of Children", title = "Children Who Did Not Receive Needed Care by Poverty Level", 
 caption = "Based on National Survey of Children's Health, 2019-2020", fill = "Poverty Level") +
  coord_flip() +
 theme_minimal()

ggplotly(CarePovChart, tooltip = "text")
```


### By Race 

```{r care by race}
CareRaceChart <- childhealthdata %>%
 filter(!(ForegoneCare %in% "Unknown")) %>%
  filter(!(RaceCat %in% "Unknown")) %>%
 ggplot() +
 aes(x = RaceCat, fill = ForegoneCare, text = paste("Children: ", ..count.., "<br> Care: ", fill)) +
 geom_bar(position = "fill") +
 scale_fill_viridis_d(option = "viridis", 
 direction = 1) +
  scale_y_continuous(labels = scales::percent) +
 labs(x = "", y = "Proportion of Children", title = "Children Who Did Not Receive Needed Care by Race", 
 caption = "Based on National Survey of Children's Health, 2019-2020", fill = "Accessing Care") +
  coord_flip() +
 theme_minimal()

ggplotly(CareRaceChart, tooltip = "text")
```


Outcomes and Wellbeing {data-orientation=columns}
=====================================

Column {data-width=450} 
--------------------------------------------------------------
### Overview 
Adverse Childhood Experiences, known as "ACEs", are potentially traumatic events that occur in childhood (ages 0-17). The events, included here, are not necessarily traumatic but have the potential to be, especially exposure to multiple. Here we look at ACEs in relation to overall health status (assessed by asking "In general, how would you describe this child's health?") & "Flourishing" status as assessed by responses to a set of 3 items (3 being ideal). 
We see ACES correlated with worse health status, with worse  “flourishing” status and that the sum of ACEs across racial categories is disparate. That is, **ACEs correlate with worse outcomes and we see some Races disproportionately impacted.**

Column  {data-width=850} 
-------------------------------------------------------------

### ACEs by Racial Category
```{r ACEs and Race}
ACEandRACE <- childhealthdata %>%
  filter(!(RaceCat %in% "Unknown")) %>%
  filter(!(ACEs %in% "Missing")) %>%
 ggplot() +
 aes(x = RaceCat, fill = ACEs, text = paste("Children:", ..count..)) +
 geom_bar(position = "fill") +
  coord_flip() +
 scale_fill_viridis_d(option = "viridis", 
 direction = 1) +
  scale_y_continuous(labels = scales::percent) +
 theme_minimal()

ggplotly(ACEandRACE, tooltip = c("text", "ACEs")) %>% layout(title = "Adverse Childhood Experiences (ACE) by Race in the US", font=list(size = 10), xaxis = list(title = 'Proportion of Children', size = 9), yaxis = list(title = 'Race', size = 9), legend = list(title = 'Number of ACEs', size = 9))


```

### ACEs and Flourishing Status
```{r ACE and Flourish}


FlrshAcePlot <- childhealthdata %>% 
  filter(!(FlourishStatus6to17 %in% "Unknown")) %>%
  filter(!(FlourishStatus6to17 %in% "NA, Under 5")) %>%
  filter(!(ACEs %in% "Missing")) %>%
  filter(!(ACEs %in% "Unknown")) %>%
ggplot() +
 aes(x = FlourishStatus6to17, fill = ACEs, text = paste("Children:", ..count..)) +
 geom_bar(stat="count", position = "fill") +
  coord_flip() +
  scale_fill_viridis_d(option = "viridis", 
 direction = 1) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Flourishing Items Met", y = "Proportion of Children", fill = "Number of ACEs") +
 theme_minimal()


ggplotly(FlrshAcePlot, tooltip = c("text", "ACEs")) %>% layout(title = "Flourishing Status & Adverse Childhood Experiences (ACE)", font=list(size = 10), xaxis = list(title = 'Proportion of Children', size = 9), yaxis = list(title = 'Flourishing Items', size = 9), legend = list(title = 'Number of ACEs', size = 9))
```

Column {data-width=450} 
-------------------------------------------------------------

### ACE List

**Possible ACEs** <br>
1. Hard to cover basics on family's income <br>
2. Parent or guardian divorced or separated <br>
3. Parent or guardian died <br>
4. Parent or guardian served time in jail <br>
5. Saw or heard parents or adults slap, hit, kick punch one another in the home <br>
6. Was a victim of violence or witnessed violence in their neighborhood <br>
7. Lived with anyone who was mentally ill, suicidal, or severely depressed <br>
8. Lived with anyone who had a problem with alcohol or drugs <br>
9. Treated or judged unfairly due to race/ethnicity <br>
10. Treated or judged unfairly due to sexual orientation or gender identity 
